Wrattler: \textnormal{A platform for AI-assisted data science}

### 2018-02-28 report/paper.tex
assistants section; add lots to intro

\section{Introduction}
Data science is an iterative, exploratory process that requires a collaboration between a 
computer system and a human. A computer can provide advice based on statistical analysis of the 
data and discover hidden structures or corner cases, but only a human can decide what those mean
and decide how to handle them in data science scripts.

A data science environment of the future needs to support an interaction model that keeps the 
human in the loop, allows an efficient human-computer interaction and can be extended with new 
AI asistants that provide advice about data. Such advice needs to be explainable to the human user,
as well as machine-readable by other components of the system. In this report, we present Wrattler, 
a system that is:

\paragraph{Interactive.}
Wrattler enables an efficient interaction by bringing computation closer to the human.
Notebooks run in the browser, cache partial results of computations and provide previews
of script results on-the-fly during development.

\paragraph{Reproducible.} 
Wrattler separates the task of running scripts from the task of managing state. 
This is handled by a data store, which tracks the provenance and semantics of data, supports 
versioning and keeps the history, making the data analyses fully reproducible.

\paragraph{Polyglot.}
Multiple languages can be used in a single notebook and share data via the data store.
Analysts can use R and Python, but also interactive languages for data exploration
that run in the browser and provide live previews.

\paragraph{Smart.}
Wrattler serves as a platform for AI assistants that use machine learning to provide suggestions 
about data. Such AI assistants connect to the data store to infer types and meaning of data, provide 
help with data cleaning and joining, but also help data exploration by finding typical and atypical 
data points and automatically visualizing data.

@@

{\color{blue}
An AI assistant guides data analysts through a single data collection, integration, preparation, 
analytics, or reporting task. Typically analysts will interact with many separate assistants over 
the course of an analysis, each of which specialize in a single kind of task.
}
\subsection{Architecture of AI assistants}

{\color{blue}
AI assistants have access to information in the Wrattler data store and can use two separate
aspects for their guidance:
%
\begin{itemize}
\item[--] AI assistants can access all data stored in the data
store. This includes imported external data files, such as raw text or raw HTML content from which
the analyst wants to extract data, as well as structured data frames imported from CSV files or
produced as a result of previous computations. Moreover, an AI assistant can also access multiple
versions of a file.
\vspace{0.25em}

The assistants should be self-contained in the sense that they interact only through the 
data store. This restriction is designed to encourage the Unix "do one thing well" design for 
assistants.

There are two important design aspects of AI assistants. First, they always produce code, rather
than being a black-box that transforms data. Second, they are designed to be used interactively.

@@

The output snippet should be readable, and something that a data analyst can look at to verify 
that the transformation accomplishes what they had wanted. As far as possible, the snippet 
should be "reusable" in a sense that it could be copy-pasted into another script that was 
processing a similar dataset and would still work. You can think of a snippet as something 
that a person might write in a single cell of a Jupyter notebook.
}

\subsection{Types of AI assistants}
{\color{blue}
AI assistants can be imagined for every step of the data analytics process. We attempt a fairly c
omprehensive list of the data analytics tasks for which can imagine targetting assistants. Then, 
for each task, we list a few examples of specific agents. Even though the list of agents is 
incomplete, it is still too long to implement fully in the current project.

\paragraph{Data collection.} Assistants can search the data store, as well as external repositories
to recommend data sets that are relevant to the task at hand and offer to import these. For example,
when analysing data about different countries, more information about countries can be offered.

\paragraph{Data integration.} Assistants can identify pairs of variables across multiple data 
sources that are likely to be coding the same variable.

\paragraph{Schema inference.} Inferring types of variables and doing this probabilistically.

\paragraph{Data parsing.}
Quick tool for specifying a scraper for web pages (related to wrapper induction, Sumit Gulwani 
work, possibly this paper Robust and Noise Resistant Wrapper Induction
Assistant to automatically infer format of csv files
Assistants to extract tables from pdf's, html
Identify which values are likely to be "special", like missing data markers

\paragraph{Data cleaning}
Identifying values that seem to be out of range, e.g. missing value indicators
Identifying possible inconsistent coding of strings
Identifying potential functional dependencies (1:1 correspondences) and returning violations
Manual cleaning of subset of data (a la Sample Clean)
Indentifying rows that seem "jointly" unusual, i.e., that seem not to conform to statistical dependencies among the columns
Identifying distinct records that may refer to the same entity (record linkage)

\paragraph{Data monitoring.}
Data diff: Returning a report of the differences between two data sets
Monitoring classifier to detect shifts in its input distribution
Detecting change points and covariate shift
Montioring classifier to suggest when/how often new data points shoud be labelled for monitoring

\paragraph{Exploratory visualization.}
Summarizing most representative rows of a data table
Automatically styling scatterplots and bar charts

\paragraph{Performing analytics.}
Which classification methods should I use? e.g., by search through similar problems on Kaggle and Open ML (a la AutoML)
Simiarly for feature selection methods and so on
Hyperparameter exploration

\paragraph{Debugging and interpreting predictive analytics.}
Debugging predictive analytics, e.g., deciding whether and how to add more features
Error analysis assistant: Summarize the output on validation data
Error analysis based on visualization of classifier score (see work of Saleema Amershi)
Tools for explaining the predictions of a model on validation data (a la LIME) * Research directions: Improved model families for the interpretable explanation. Decision lists on marginal distribution rather than joint
"Training set blame": Given an error in the validation data, what examples from the training set caused me to make that error?
Explaining differences between predictions: why was instance 24601 treated differently than 24675?
}
\subsection{Possible AI assistants}

@@

*Research questions:* First, are there research questions here, or is this so simple
that it  is part of the infrastructure?  If $k$-medoids is so simple for this,
why doesn't everyone also do it? How to evaluate whether the summary is good?
Maybe there are examples of data sets where medoids is not good.
For example, in the Karpathy ICML example, medoids would not make sense,
listing the institutions that have the most papers is most informative, because
it's a "long tail" type of column. How do you tell "long tail" data from "just show
the clusters" type data? Perhaps a model-based framework could distinguish?
Maybe you want to summarize the ways in which two data sets differ?
Is the best summary the cluster centroids or the "top k" along some value? 
* *User interaction:* I'm not sure what interactions are enecessary.
Users could ask for more rows, mark two items as similar, or mark an item as uninteresting.
Or perhaps better would be to allow drill down, i.e., to make it easy to explore
the clusters represented by each example.
}

@@

\end{document}

>>>>

\section{Introduction}
Data science is an iterative, exploratory process that requires a collaboration between a 
computer system and a human. A computer can provide advice based on statistical analysis of the 
data and discover hidden structures or corner cases, but only a human can decide what those mean
and decide how to handle them in data science scripts. Data science is often cited as an expensive
and time consuming task, especially due the costs of data cleaning and data wrangling.
We propose four fundamental reasons why
practical data science is expensive:

\paragraph{Big data is big,} 
so the analyst doesn't understand it all.*
Even if a data set is small enough fits on one computer, it's still too large to fit in 
an analyst's working memory.
But this means every analysis is huanted
by the spectre of lurking, potentially
unknown data quality  issues.
This also makes it more difficult to do data fusion, because there may be corner cases that make it more difficult to join two disparate data sources than expected.

\paragraph{The double Anna Karenina principle.}
Not only is every dirty data set  dirty in its own way, but \emph{pace} Tolstoy, every clean data set is clean in its own way as well. 
"Data" is such an abstract concept that specific integrity conditions to characterize whether data is dirty, and potentially even the most appropriate formalism for integrity conditions,  differ dramatically across
the vast array of disparate use cases of data science,
ranging form relational data describing the customers of a country, time series data describing sensor data in an internet of things platform, and huge datasets of satellite imagery of the earth over a multi-year time scale.

\paragraph{Death by a thousand cuts.} 
Often data transformation and processing steps are individually  very simple. But  there may need to be lots of them, and
because big data is big,
an analyst never knows if she has found them all.

\paragraph{Feedback cycles everywhere.} Data science is 
not a pipeline but a connected mess of epicycles. This is because every step 
in a data analysis actually teaches 
the analyst more about the data and the problem, which might require rehtinking the earlier steps. 
For example, there might be data quality issues that 
are not uncovered until the analyst
investigates the output of a regression model.

\vspace{1em}
To meet these challenges, we present Wrattler,
a new type of system for data science
that aims to transform the process
of data analysis.
Wrattler combines the interactive and literal programming paradigms
of notebook systems such as Jupyter
with new advances in AI systems for data wrangling and in provenance.
The main design principles in Wrattler are:

\paragraph{Interactive.} Interactivity
is a necessary because ``big data is big'', so the analyst learns about 
the data set and the problem as she explores it.
Wrattler enables an efficient interaction by bringing computation closer to the human.
Notebooks run in the browser, cache partial results of computations and provide previews
of script results on-the-fly during development.

\paragraph{Reproducible.} 
Data analyses must be reproducible
because of feedback cycles. As the
analyst learns more about the problem,
this may uncover data cleaning or preparation issues that require
redesigning and rerunning the analysis.
Wrattler separates the task of running scripts from the task of managing state. 
This is handled by a data store, which tracks the provenance and semantics of data, supports 
versioning and keeps the history, making the data analyses fully reproducible.

\paragraph{Polyglot.}
Modern data science naturally draws
on many competing languages and libraries, such as R and Scipy. As a side effect
of our interactive, reproducible design,
we obtain nearly for free the ability
to support polyglot data analyses.
Multiple languages can be used in a single notebook and share data via the data store.
Analysts can use R and Python, but also interactive languages for data exploration
that run in the browser and provide live previews.

\paragraph{Smart.}
AI can examine and find patterns
in big data where a human cannot,
again aiming at the problem that
big data is big.
AI can be used to direct the analyst's attention and to generalize  decisions about data transformation
to new data that that the analyst
hasn't seen.
Wrattler serves as a platform for AI assistants that use machine learning to provide suggestions 
about data. Such AI assistants connect to the data store to infer types and meaning of data, provide 
help with data cleaning and joining, but also help data exploration by finding typical and atypical 
data points and automatically visualizing data.

@@

{
An AI assistant is a component
of the system that 
guides data analysts through a single data collection, integration, preparation, 
analytics, or reporting task.
For example, an assistant might
specialize in identifying suspicious
values, such as "-1" in a column labelled
"heart rate", which might indicate
that certain data was not collected.
Assistants are interactive, in that
they propose a transformation
of the data to the analyst, based on combinations
of statistical and symbolic AI analysis,
and then refine the transformation
based on feedback from the analyst.
Typically analysts will interact with many separate assistants over 
the course of an analysis, each of which specialize in a single kind of task.
}
\subsection{Architecture of AI assistants}

There are two important design aspects of AI assistants, which may seem 
contradictory.
 First, they must be interactive, following
 the first principle behind Wrattler, but
 they must also be reproducible.
We square this circle by using the notebook
system as a reproducible interlingua. 
The output of an AI assistant is always code, rather
than being a black-box that transforms data,
and it is this code that is agreed between the AI and the analyst during the interaction.
Thus the interaction is viewed as a ``design phase''
that is not reproduced on new data.

Assistants are therefore based on two different ideas:

@@

The output snippet should be readable, and something that a data analyst can look at to verify 
that the transformation accomplishes what they had wanted. As far as possible, the snippet 
should be "reusable" in a sense that it could be copy-pasted into another script that was 
processing a similar dataset and would still work. You can think of a snippet as something 
that a person might write in a single cell of a Jupyter notebook.

\vspace{1em}
AI assistants have access to information in the Wrattler data store and can use two separate
aspects for their guidance:
%
\begin{itemize}
\item[--] AI assistants can access all data stored in the data
store. This includes imported external data files, such as raw text or raw HTML content from which
the analyst wants to extract data, as well as structured data frames imported from CSV files or
produced as a result of previous computations. Moreover, an AI assistant can also access multiple
versions of a file.
\vspace{0.25em}

The assistants should be self-contained in the sense that they interact with each other
only through the 
data store. This restriction is necessary in order
for provenance tracking and reproducibility.

\subsection{The Assistant Landscape}
{

AI assistants can be imagined for every step of the data analytics process. 
To give an idea of the breadth of this framework,
we present a broad list of the data analytics tasks which have potential for assistant automation.
This is an intentionally broad list, more so than
could be fully explored in even a large research project, but gives a sense of the breadth of 
potential use cases of our framework.

\paragraph{Data collection.} Assistants can search the data store, as well as external repositories
to recommend data sets that are relevant to the task at hand and offer to import these. For example,
when analysing data about different countries, 
the assistant could suggest joining in public
demographic information from DBPedia.

\paragraph{Data integration} is a problem as common
as it is difficult. We envisage assistants
that combine logical and statistical reasoning to assist
with data  integration. A simple but useful
example of a tool in this space is to
 identifying pairs of variables across multiple data 
sources that are likely to be daily temperatures.
More sophisticated examples could include and extend the
most cutting edge techniques in the data fusion and
semantic technologies literatures.

\paragraph{Schema inference.} A sad fact of data
science is that data sets are often not as well documented
as we might like, even to the minimal level of
documentation of the types of each column.
There is an interesting opportunity to
develop machine learning methods that inferring types of variables from their values, e.g.,
by noticing that integer values between 0 and 30 in 
the United Kingdom might describe an outdoor temperature, especially if other weather-related
variables are detected in the data set.

\paragraph{Data parsing.}
A common task in data analysis is to convert
data that is ``almost'' in structured form,
such as HTML tables, into a relational table.
Assistants in this space include
tools specifying a scraper for web pages, related
to wrapper induction \cite{Kushmerick1997WrapperIF},
for inferring format of CSV files (while simple
conceptually, they have no standard format), and
for extracting tables from html and pdf files
\cite{pinto03table}.

\paragraph{Data cleaning} is a well studied
problem in the databases and data mining literature
\cite{abedjan2016detecting,ilyas2015},
but there are still a huge number of issues that have been unexplored.
For example, we envisage assistants for
Identifying values that seem to be out of range, e.g. missing value indicators;
Identifying possible inconsistent coding of strings,
such as different representations of acrynoms;
indentifying rows that seem "jointly" unusual, i.e., that seem not to conform to statistical dependencies among the columns; and
identifying distinct records that may refer to the same entity (record linkage).

\paragraph{Data monitoring.} Once an analysis
is completed, it may result in a machine learning
model which will then be deployed on a data stream.
Once a classifier is deployed, it must be monitored 
to detect shifts in its input distribution,
such as change points, covariate shift, and
concept drift. This includes the problem 
to suggesting when/how often new data points should be labelled for monitoring.
Even for data analyses that do not involve
deploying a ``production model'', there is
often a need to run an analysis periodically on updated data, such as re-running last quarter's
analysis on this quarter's data.
In such cases, we might believe that the two data
sets come from approximately the same distribution
and format, so we can exploit
divergences from this assumption to indicate
potential issues with data quality.
As part of the Wrattler effort, we have
introduced the \emph{data diff} problem \cite{datadiff},
which aims to improve the process of data wrangling by 
returning a report of the differences between two data sets.

\paragraph{Exploratory visualization} to summarize
data sets, both the target data of the analysis,
and the results of transformations that the analyst
creates. Potential assistants here include ones
to summarize most representative rows of a data table,
and automatically choosing appropriate interactive
visualizations, such as scatterplots and bar charts,
with styling decisions such as axes automatically
chosen to be most informative.

\paragraph{Performing analytics.}
An increasing amount of attention is being
given to the AutoML challenge \cite{guyon_review_2016}, which focuses
on automatically choosing a classification
methods, feature representations and hyperparameters
for a given classification task, without the need for human intervention.

\paragraph{Debugging and interpreting predictive analytics.}
Every time a new machine learning model is trained,
the first question an analyst has is often
why its accuracy is not better. But
debugging predictive analytics is notoriously difficult, and the remedies indirect and expensive,
 e.g., labelling more data, adding more features,
 using a more complex model. Debugging and improving
 models is a rich area for potential assistants.
 There is potentially low hanging fruit available
 in how to aid the tasks of
error analysis, exploring the
predictions of the model 
on validation data. Some early work in this direction is
\cite{saleema}.
There is a rich and growing literature on
interpreting and explaining the predictions
of a model \cite{lipton:mythos,doshi-velez17,ribiero2016lime,darksight}.
Two potential directions from data are
debugging by using
"training set blame": Given an error in the validation data, what examples from the training set caused me to make that error? This might build on the work of \cite{percy}.
Another potential avenue is explaining
differences between predictions: why was instance 42 treated differently than instance 24601, even though they appear highly similar to a human?
}

\vspace*{1em}

%%cs possibly we don't need to go into detail
%% in this version of the report? Or we might
%% decide to add this back.
\begin{comment}
\subsection{Possible AI assistants}

Designing an AI assistant requires an intriguing
combination of data science, AI, and user interaction
thinking.
As an example of the types of design considerations
that arise, we go into more detail about the
design issues that arise when considering
three examples of the broader list of agents
in the previous section.

@@

*Research questions:* First, are there research questions here, or is this so simple
that it  is part of the infrastructure?  If $k$-medoids is so simple for this,
why doesn't everyone also do it? How to evaluate whether the summary is good?
Maybe there are examples of data sets where medoids is not good.
For example, in the Karpathy ICML example, medoids would not make sense,
listing the institutions that have the most papers is most informative, because
it's a "long tail" type of column. How do you tell "long tail" data from "just show
the clusters" type data? Perhaps a model-based framework could distinguish?
Maybe you want to summarize the ways in which two data sets differ?
Is the best summary the cluster centroids or the "top k" along some value? 
* *User interaction:* I'm not sure what interactions are enecessary.
Users could ask for more rows, mark two items as similar, or mark an item as uninteresting.
Or perhaps better would be to allow drill down, i.e., to make it easy to explore
the clusters represented by each example.
}
\end{comment}

@@

\bibliographystyle{plain}
\bibliography{paper}
\end{document}
